---
title: "Shoulder instability transcriptome analysis"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/shoulder-instability-osteroarthritis

# Introduction

```{r,packages}

suppressPackageStartupMessages({
    library("zoo")
    library("tidyverse")
    library("reshape2")
    library("DESeq2")
    library("gplots")
    library("fgsea")
    library("MASS")
    library("mitch")
    library("eulerr")
    library("limma")
    library("topconfects")
    library("kableExtra")
})

```

## Import read counts

```{r,importdata}

tmp <- read.table("3col.tsv.gz",header=F)
x <- as.matrix(acast(tmp, V2~V1, value.var="V3", fun.aggregate = sum))
x <- as.data.frame(x)
accession <- sapply((strsplit(rownames(x),"\\|")),"[[",2)
symbol<-sapply((strsplit(rownames(x),"\\|")),"[[",6)
x$geneid <- paste(accession,symbol)
xx <- aggregate(. ~ geneid,x,sum)
rownames(xx) <- xx$geneid
colnames <- gsub("T0R","T0",colnames(xx))
xx$geneid = NULL
xx <- round(xx)

head(xx)

```

## QC analysis

Here I'll look at a few different quality control measures.

```{r,qc1,fig.height=7,fig.width=7}

par(mar=c(5,8,3,1))
barplot(colSums(xx),horiz=TRUE,las=1,xlab="num reads")

sums <- colSums(xx)
sums <- sums[order(sums)]
barplot(sums,horiz=TRUE,las=1,xlab="num reads")
abline(v=15000000,col="red")

```

## Sample sheet

```{r,samplesheet1}

URL="https://raw.githubusercontent.com/markziemann/shoulder-instability-osteroarthritis/main/OA_samplesheet.tsv"
ss <- read.table(URL,header=TRUE)
ss$col <- as.numeric(factor(ss$TissueType))

```

## MDS plot

Multidimensional scaling plot to show the variation between samples, very similar to PCA.

```{r,mds,fig.height=7,fig.width=7}

colvec <- c("orange2","pink","purple1","yellow","lightblue")
cols <- rep(colvec,6)

plot(cmdscale(dist(t(xx))), xlab="Coordinate 1", ylab="Coordinate 2", 
  type = "p",bty="n",pch=19, cex=4, col=cols)
text(cmdscale(dist(t(xx))), labels=ss$ParticipantID )

legend("topright", inset=0.1, title="tissue type", c("B","C","F","M","S"),
  pch=19,col=colvec, cex=1.5)

```

## Correlation heatmap

```{r,cor,fig.height=7,fig.width=7}

heatmap.2(cor(xx),trace="n",main="Pearson correlation heatmap")

```

## Findings

There were 4 samples with fewer than 15M reads and this might impact data quality: 4002-42S, 4006-46B, 4005-42F and 4006-46C.

MDS plot shows samples clustered by tissue type as expected.
For example M samples to the right, F samples in the centre, C samples in the lower right and S samples between C and F.
However the B samples were somewhat not clustered, which is something to look into.

The result of the MDS is confirmed in the correlation heatmap.

## Session information

For reproducibility.

```{r,sessioninfo}

sessionInfo()

```
