---
title: "Shoulder transcriptome analysis: rotator cuff repair"
author: "Mark Ziemann"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
    fig_width: 7
    fig_height: 7
theme: cosmo
---

Source: https://github.com/markziemann/shoulder-instability-osteroarthritis

# Introduction

Predict gene expression profile in capsular tissue biopsies of patients undergoing rotator
cuff repair (AD-CAB 3001 to AD-CAB 3026) using capsular tissue biopsies in patients undergoing
shoulder replacement for primary OA (AD-CAB 4001-4020, 'Yes' column 10 'Primary OA') or
Rotator cuff arthropathy (AD-CAB 4001-4020, 'Yes' column 11 'Cuff arthropathy')?
(adjusting for age and sex also).

* First step, do a PCA of these 3 groups (RCR, OA, RCA) and comparison expression profiles.

```{r,packages}

suppressPackageStartupMessages({
    library("R.utils")
    library("zoo")
    library("tidyverse")
    library("reshape2")
    library("DESeq2")
    library("gplots")
    library("fgsea")
    library("MASS")
    library("mitch")
    library("eulerr")
    library("limma")
    library("topconfects")
    library("kableExtra")
})

```

## Import count data and sample sheet

caps_rcr= capsule rotator cuff repair.

caps_oa= capsule of primary OA

caps_rca= capsule rotator cuff arthropathy

caps_all= includes all three above groups.

```{r,samplesheet}

x <- readRDS("counts.rds")

ss <- read.table("samplesheets/pheno_data.tsv",header=TRUE)
head(ss,3)

sapply(strsplit(gsub("-",".",gsub("_L001","|",gsub("_R1","|",ss$fastq))),"\\|"),"[[",1) %in% colnames(x)
ss$data <- sapply(strsplit(gsub("-",".",gsub("_L001","|",gsub("_R1","|",ss$fastq))),"\\|"),"[[",1) 

rownames(ss)<-ss$data

caps <- subset(ss,Tissue=="capsule")
head(caps,2)
dim(caps)

caps_rcr <- caps[grep("_30",caps$Participant_ID),]
head(caps_rcr,3)
dim(caps_rcr)
caps_rcr$group="rcr"
caps_rcr %>% kbl(caption = "Capsule samples of rotator cuff repair group") %>% kable_paper("hover", full_width = F)


caps_oa <- subset(caps,Primary_OA=="Yes")
head(caps_oa,3)
dim(caps_oa)
caps_oa$group="oa"
caps_oa %>% kbl(caption = "Capsule samples of primary OA group") %>% kable_paper("hover", full_width = F)

caps_rca <- subset(caps,Cuff_Arthropathy=="Yes")
head(caps_rca,3)
dim(caps_rca)
caps_rca$group="rca"
caps_rca %>% kbl(caption = "Capsule samples of rotator cuff arthropathy group") %>% kable_paper("hover", full_width = F)

caps_all <- rbind(caps_rcr,caps_oa,caps_rca)

xcaps <- x[,which(colnames(x) %in% rownames(caps_all) )]

dim(xcaps)
dim(caps_all) # matches OK

caps_all %>% kbl(caption = "Capsule samples") %>% kable_paper("hover", full_width = F)

```

## QC analysis

First to look at the number of reads per data set.

```{r,qc1,fig.height=7,fig.width=7}

par(mar=c(5,8,3,1))
barplot(colSums(xcaps),horiz=TRUE,las=1,xlab="Capsule reads")

sums <- colSums(xcaps)
sums <- sums[order(sums)]
barplot(sums,horiz=TRUE,las=1,xlab="Capsule reads")
abline(v=15000000,col="red")

```

## MDS plot

Multidimensional scaling plot to show the variation between all samples, very similar to PCA.

```{r,mds,fig.height=7,fig.width=7}

mds <- cmdscale(dist(t(xcaps)))

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=4 ,col="gray")
text(mds, labels=rownames(mds) )

rcr1 <- as.numeric(colnames(xcaps) %in% rownames(caps_rcr))*1
oa2 <- as.numeric(colnames(xcaps) %in% rownames(caps_oa))*2
rca3 <- as.numeric(colnames(xcaps) %in% rownames(caps_rca))*3

mycols <- gsub("3","gray",gsub("2","pink",gsub("1","lightblue",as.character(rcr1+oa2+rca3))))

plot(mds, xlab="Coordinate 1", ylab="Coordinate 2",
  type = "p",bty="n",pch=19, cex=4 ,col=mycols)
text(mds, labels=rownames(mds) )
mtext("RCR=lightblue, OA=pink, RCA=grey")

```

## Primary OA vs Rotator cuff arthropathy

```{r,de1}

ss1 <- rbind(caps_oa,caps_rca)
ss1 <- ss1[order(rownames(ss1)),]

x1 <- xcaps[,colnames(xcaps) %in% rownames(ss1)]
x1 <- x1[,order(colnames(x1)),]

colnames(x1) == rownames(ss1)

x1f <- x1[which(rowMeans(x1)>10),]
dim(x1)
dim(x1f)

dds <- DESeqDataSetFromMatrix(countData = x1f , colData = ss1, design = ~ Age +Sex + Cuff_Arthropathy )

res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression differences between OA (ctrl) and Rotator cuff arthropathy (case)") %>%
  kable_paper("hover", full_width = F)
dge1 <- dge
write.table(dge1,file="rc1.tsv",quote=FALSE,sep='\t')

```

## Primary OA vs Rotator cuff repair

```{r,de2}

ss2 <- rbind(caps_oa,caps_rcr)
ss2 <- ss2[order(rownames(ss2)),]

x2 <- xcaps[,colnames(xcaps) %in% rownames(ss2)]
x2 <- x2[,order(colnames(x2)),]

colnames(x2) == rownames(ss2)

x2f <- x2[which(rowMeans(x2)>10),]
dim(x2)
dim(x2f)

dds <- DESeqDataSetFromMatrix(countData = x2f , colData = ss2, design = ~ Age +Sex + group )

res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression differences between OA (ctrl) and Rotator cuff repair (case)") %>%
  kable_paper("hover", full_width = F)
dge2 <- dge
write.table(dge2,file="rc2.tsv",quote=FALSE,sep='\t')

```

## Rotator cuff repair vs Rotator cuff arthropathy

```{r,de3}

ss3 <- rbind(caps_rcr,caps_rca)
ss3 <- ss3[order(rownames(ss3)),]

x3 <- xcaps[,colnames(xcaps) %in% rownames(ss3)]
x3 <- x3[,order(colnames(x3)),]

colnames(x3) == rownames(ss3)

x3f <- x3[which(rowMeans(x3)>10),]
dim(x3)
dim(x3f)

ss3$group <- factor(ss3$group,levels=c("rcr","rca"))
dds <- DESeqDataSetFromMatrix(countData = x3f , colData = ss3, design = ~ Age +Sex + group )

res <- DESeq(dds)
z <- results(res)
vsd <- vst(dds, blind=FALSE)
zz <- cbind(as.data.frame(z),assay(vsd))
dge <- as.data.frame(zz[order(zz$pvalue),])
head(dge,20) %>%
  kbl(caption = "Top gene expression differences between OA (ctrl) and Rotator cuff repair (case)") %>%
  kable_paper("hover", full_width = F)
dge2 <- dge
write.table(dge2,file="rc2.tsv",quote=FALSE,sep='\t')

```
## Session information

For reproducibility.

```{r,sessioninfo}

sessionInfo()

```
